Binary Latent Representations for Efficient Ranking

### 2017-06-20 paper/paper.tex
Change names to use 1M dataset.

\begin{abstract}
Large-scale recommender systems often face severe latency and storage constraints at prediction time. These are particularly acute when the number of items that could be recommended is large, and calculating predictions for the full set is computationally intensive. To relax these constraints, we train recommendation models that use binary rather than real-valued user and item representations, and show that they offer a range of attractive speed-accuracy trade-offs. In particular, we demonstrate 13 times faster prediction at the expense of a modest 4\% loss in ranking accuracy on the Movielens 100K dataset.
\end{abstract}

@@

To alleviate these constraints in both online and offline settings, we propose the use of binary item and user representations in learning-to-rank matrix factorization models. Following the approach of \citet{rastegari2016xnor}, we show that we can estimate binary user and item representations that achieve comparable ranking accuracy to standard learning-to-rank approaches while being several times faster to score, requiring a fraction of the memory, and being only marginally more complex to estimate. In MovieLens 100k experiments, our binary representations are within 4\% of the accuracy of the real-valued model, but are over 13 times faster to score and take only 6\% of the memory.

@@

\section{Experiments}
\begin{table*}[htbp]
\begin{threeparttable}
\caption{Movielens 100K results}
\label{tb:results}
\centering
\begin{tabular}{rrrrrrrr}
\toprule
Dimension &   MRR & Binary MRR & MRR ratio\tnote{1} &   PPMS\tnote{2} & Binary PPMS & PPMS ratio & Memory use ratio \\
\midrule
       32 & 0.096 &      0.088 &     0.919 & 89,299 &     245,644 &      2.751 &            0.091 \\
       64 & 0.098 &      0.096 &     0.973 & 47,565 &     209,486 &      4.404 &            0.062 \\
      128 & 0.108 &      0.091 &     0.843 & 20,502 &     156,404 &      7.629 &            0.047 \\
      256 & 0.109 &      0.097 &     0.884 & 10,631 &     166,500 &     15.661 &            0.039 \\
      512 & 0.112 &      0.105 &     0.938 &   5391 &     110,418 &     20.479 &            0.035 \\
     1024 & 0.114 &      0.102 &     0.897 &   2689 &      69,349 &     25.786 &            0.033 \\
     2048 & 0.111 &      0.109 &     0.984 &   1352 &      34,826 &     25.745 &            0.032 \\
     4096 & 0.111 &      0.107 &     0.961 &    666 &      15,907 &     23.856 &            0.032 \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small{
\item[1] Ratio of binary model MRR to real-valued model MRR
\item[2] Predictions Per Millisecond
}
\end{tablenotes}
\end{threeparttable}
\end{table*}
To assess the accuracy-speed trade-offs enabled by the XNOR approach, we conduct an experiment on the Movielens 100K dataset \citep{harper2016movielens}. The dataset contains 100 thousand ratings from 943 users on 1682 movies. Because the computational speed improvements offered by the binary representations are linear in the size of the catalogue, the results on this dataset should be indicative of results that can be obtained on larger datasets. At the same time, the dataset's small size enables us to easily test hyperparameter configurations and estimate models with high latent dimensions.

\subsection{Experimental setup}
We randomly divide the dataset into a training, test, and validation set. In order to build a picture of the accuracy-speed trade-offs afforded by real-valued and binary latent models, we explicitly build models for between 32 and 4096 latent dimensions. For every latent dimensionality, we conduct a random search over the hyperparameter space, and pick optimal learning rates and minibatch sizes for each algorithm based on their performance on the test set. The final results are obtained from ranking interactions from the validation set. We use mean reciprocal rank (MRR) as a measure of ranking quality.

@@

\subsection{Results}
\label{sec:results}
Table \ref{tb:results} summarises the results of the Movielens 100K experiment. Each row presents the best MRR (mean reciprocal rank) results for both real-valued and binary models of a given dimensionality. MRR ratio denotes the fraction of the real-valued MRR the binary model achieves for that dimensionality; PPMS (predictions per millisecond) ratio denotes how many more predictions the binary model can compute per millisecond.

>>>>

\begin{abstract}
Large-scale recommender systems often face severe latency and storage constraints at prediction time. These are particularly acute when the number of items that could be recommended is large, and calculating predictions for the full set is computationally intensive. To relax these constraints, we train recommendation models that use binary rather than real-valued user and item representations, and show that they offer a range of attractive speed-accuracy trade-offs. In particular, we demonstrate 13 times faster prediction at the expense of a modest 4\% loss in ranking accuracy on the Movielens 1M dataset.
\end{abstract}

@@

To alleviate these constraints in both online and offline settings, we propose the use of binary item and user representations in learning-to-rank matrix factorization models. Following the approach of \citet{rastegari2016xnor}, we show that we can estimate binary user and item representations that achieve comparable ranking accuracy to standard learning-to-rank approaches while being several times faster to score, requiring a fraction of the memory, and being only marginally more complex to estimate. In MovieLens 1M experiments, our binary representations are within 4\% of the accuracy of the real-valued model, but are over 13 times faster to score and take only 6\% of the memory.

@@

\section{Experiments}
\begin{table*}[htbp]
\begin{threeparttable}
\caption{Movielens 1M results}
\label{tb:results}
\centering
\begin{tabular}{rrrrrrrr}
\toprule
Dimension &   MRR & Binary MRR & MRR ratio\tnote{1} &   PPMS\tnote{2} & Binary PPMS & PPMS ratio & Memory use ratio \\
\midrule
       32 & 0.096 &      0.088 &     0.919 & 89,299 &     245,644 &      2.751 &            0.091 \\
       64 & 0.098 &      0.096 &     0.973 & 47,565 &     209,486 &      4.404 &            0.062 \\
      128 & 0.108 &      0.091 &     0.843 & 20,502 &     156,404 &      7.629 &            0.047 \\
      256 & 0.109 &      0.097 &     0.884 & 10,631 &     166,500 &     15.661 &            0.039 \\
      512 & 0.112 &      0.105 &     0.938 &   5391 &     110,418 &     20.479 &            0.035 \\
     1024 & 0.114 &      0.102 &     0.897 &   2689 &      69,349 &     25.786 &            0.033 \\
     2048 & 0.111 &      0.109 &     0.984 &   1352 &      34,826 &     25.745 &            0.032 \\
     4096 & 0.111 &      0.107 &     0.961 &    666 &      15,907 &     23.856 &            0.032 \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small{
\item[1] Ratio of binary model MRR to real-valued model MRR
\item[2] Predictions Per Millisecond
}
\end{tablenotes}
\end{threeparttable}
\end{table*}
To assess the accuracy-speed trade-offs enabled by the XNOR approach, we conduct an experiment on the Movielens 1M dataset \citep{harper2016movielens}. The dataset contains 1 million ratings from 6000 users on 4000 movies. Because the computational speed improvements offered by the binary representations are linear in the size of the catalogue, the results on this dataset should be indicative of results that can be obtained on larger datasets. At the same time, the dataset's small size enables us to easily test hyperparameter configurations and estimate models with high latent dimensions.

\subsection{Experimental setup}
We randomly divide the dataset into a training, test, and validation set. In order to build a picture of the accuracy-speed trade-offs afforded by real-valued and binary latent models, we explicitly build models for between 32 and 2048 latent dimensions. For every latent dimensionality, we conduct a random search over the hyperparameter space, and pick optimal learning rates and minibatch sizes for each algorithm based on their performance on the test set. The final results are obtained from ranking interactions from the validation set. We use mean reciprocal rank (MRR) as a measure of ranking quality.

@@

\subsection{Results}
\label{sec:results}
Table \ref{tb:results} summarises the results of the Movielens 1M experiment. Each row presents the best MRR (mean reciprocal rank) results for both real-valued and binary models of a given dimensionality. MRR ratio denotes the fraction of the real-valued MRR the binary model achieves for that dimensionality; PPMS (predictions per millisecond) ratio denotes how many more predictions the binary model can compute per millisecond.
